---
title: Integrative Mendelian randomization for detecting exposure-by-group interactions using group-specific and combined summary statistics
author: "Ke Xu,Nathaniel Maydanchik, Bowei Kang, Jianhai Chen,
Qixiang Chen, Gongyao Xu, Shinya Tasaki, David A. Bennett, Lin S. Chen"
output:
  pdf_document: default
  latex_engine: xelatex
  word_document: default
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction
This vignette provides an introduction to the `int2MR` package. The R package `int2MR` implements the int2MR method for detecting both direct exposure-outcome effect within comparison group and reference group and exposure-group interaction effect.

Before installing int2MR, ensure that you have the devtools package installed. The package also requires rstan for Bayesian modeling. To install the development version of int2MR, run:
```{r}
# Load devtools package
library(devtools)

# Install the int2MR package from GitHub
install_github("Likeli-Ke/int2MR")

# Load the int2MR package
library(int2MR)
```

# Input Data Format

The `int2MR` package supports two types of GWAS summary statistics:

- **Three-sample data**: Utilizes three separate IV-to-outcome GWAS statistics.
- **Two-sample data**: Utilizes two IV-to-outcome GWAS summary statistics.

To detect the exposure-group interaction effect, it is essential that the proportion of the comparison group (denoted as rho) varies among the provided GWAS summary statistics.

## Data Requirements

Your input data should include:

- **Number of IVs** (instrumental variables)
- **Point estimates** and **squared standard errors** for:
  - IV-to-outcome effects
  - IV-to-exposure effects
- **Proportion of the comparison group (rho)** in each IV-to-outcome GWAS summary statistic.


Load the example datasets provided with the package as follows:
```{r}
data(example_3sample_data)
data(example_2sample_data)
```

For further details on the example data formats included with the package, consult the help pages.
```{r}
help(example_3sample_data)
help(example_2sample_data)
```


# Running Examples

This section demonstrates how to run `int2MR` using simulated data. Two examples are provided: one for three-sample data and one for two-sample data.

## Example 1: Three-Sample Data

In this example, we perform an analysis on simulated three-sample data. The parameters include:

- **model_type**: "3sample"
- **Prior distributions**: Inverse gamma priors with a shape and scale of 0.1
- **MCMC Settings**: 2 chains, 10000 iterations with a warm-up period of 2500 iterations, and an adapt_delta of 0.95. 

```{r}
result_3sample <- int2MR(data_list_3sample = example_3sample_data,
                 model_type = "3sample",
                 prior_inv_gamma_shape = 0.1,
                 prior_inv_gamma_scale = 0.1,
                 chains = 2, iter = 10000, warmup = 2000,
                 adapt_delta = 0.95)

# Display the results for the three-sample analysis
result_3sample$result_3sample
```

## Example 2: Two-Sample Data

This example demonstrates an analysis using simulated two-sample data. The configuration is similar to the three-sample analysis, with `model_type` set to "2sample".

```{r}
result_2sample <- int2MR(data_list_2sample = example_2sample_data,
                 model_type = "2sample",
                 prior_inv_gamma_shape = 0.1,
                 prior_inv_gamma_scale = 0.1,
                 chains = 2, iter = 10000, warmup = 2500,
                 adapt_delta = 0.95)

# Display the results for the two-sample analysis
result_2sample$result_2sample
```
