---
title: Integrative Mendelian randomization for detecting exposure-by-group interactions using group-specific and combined summary statistics
author: |
  | Ke Xu, Nathaniel Maydanchik, Bowei Kang, Jianhai Chen, Qixiang Chen, Gongyao Xu,
  | Shinya Tasaki, David A. Bennett, Lin S. Chen
output:
  pdf_document: default
  latex_engine: xelatex
  word_document: default
  html_document:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{int2MR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Introduction

This vignette provides an introduction to the `int2MR` R package. The `int2MR` package implements the int2MR method for detecting group-specific exposure-outcome effects and exposure-group interaction effects using summary statistics. 

Before installing `int2MR`, ensure that the `devtools` package is installed. `int2MR` also requires `rstan` for Bayesian modeling. To install and load the development version of `int2MR`, run:

```{r install, eval=FALSE}
# Load devtools package
suppressMessages(library(devtools))

# Install int2MR from GitHub
install_github("Likeli-Ke/int2MR")

# Load int2MR
library(int2MR)

# Load rstan
suppressMessages(library(rstan))
```

# Input Data Format

The `int2MR` package supports two types of input:

-   **Two-sample data**: For integrating two sets of IV-to-outcome GWAS summary statistics.
-   **Three-sample data**: For Integrating three sets of IV-to-outcome GWAS summary statistics.

Two-sample input data must be a list with the following components: `p` is the number of IVs; `hat_gamma` and `hat_s_gamma_sq` are the estimated IV-to-exposure effects and their variances (squared standard errors); `hat_Gamma1` and `hat_s1_sq` are the estimated IV-to-outcome effects for the first IV-to-outcome GWAS summary statistic and their variances; `hat_Gamma2` and `hat_s2_sq` are the estimated IV-to-outcome effects for the second IV-to-outcome GWAS summary statistic and their variances; `rho1` is the proportion of the comparison group in the sample of the first GWAS summary statistic; and `rho2` is the proportion of the comparison group in the sample of the second GWAS summary statistic.

```{r two-sample-data, eval=FALSE}
# Load the example two-sample data
data(example_2sample_data)
names(example_2sample_data)
```

Three-sample input data additionally contains `hat_Gamma3`, `hat_s3_sq`, and `rho3` for the third GWAS summary statistic.

```{r three-sample-data, eval=FALSE}
# Load the example three-sample data
data(example_3sample_data)
names(example_3sample_data)
```

To detect the exposure-group interaction effect, the proportion of the comparison group in the samples (rho) must vary among the provided GWAS summary statistics.

# Running Examples

This section demonstrates how to run `int2MR` using simulated data. Two examples are provided: one for two-sample data and one for three-sample data.

## Example 1: Two-Sample Data

In this example, we perform an analysis on simulated two-sample data. The parameters include:

-   **data_list_2sample**: Input data for two-sample analysis. Note that the `data_list_3sample` argument is left null
-   **model_type**: "2sample"
-   **estimation**: "sampling" for MCMC posterior mean effect estimates, or "optimizing" for optimization-based MAP effect estimates
-   **Prior distributions**: Inverse gamma priors with a shape and scale of 0.1
-   **MCMC Settings**: 2 chains, 10,000 iterations with a warm-up period of 2,500 iterations, and an adapt_delta of 0.95.

The inverse-gamma hyperparameters control the prior scale of uncorrelated pleiotropic-effect variance components. As a rule of thumb, start with `prior_inv_gamma_shape = prior_inv_gamma_scale = 0.02` for weak or sparse uncorrelated pleiotropic effects, try `0.1` for moderate effects, and check stronger effects with a small sensitivity grid such as `c(0.02, 0.05, 0.1, 0.2, 0.5)`. Report whether `est_beta` and `est_beta_int` are stable across the grid rather than relying on a single setting.

```{r two-sample-sampling, eval=FALSE}
result_2sample <- int2MR(data_list_2sample = example_2sample_data,
                 model_type = "2sample",
                 prior_inv_gamma_shape = 0.1,
                 prior_inv_gamma_scale = 0.1,
                 chains = 2, iter = 10000, warmup = 2500,
                 adapt_delta = 0.95)

# Display the results for the two-sample analysis
result_2sample$result_2sample
```
`est_beta` is the estimated group-specific exposure-outcome effect for the reference group, and `se_beta` and `pval_beta` are its standard error and p-value, respectively. `est_beta_int` is the estimated exposure-group interaction effect, and `se_beta_int` and `pval_beta_int` are its standard error and p-value, respectively. `total_effect` is the estimated group-specific exposure-outcome effect for the comparison group and `pval_total` is its respective p-value.

To estimate the effects without MCMC sampling, use `estimation = "optimizing"`. The `estimation` argument controls how `est_beta` and `est_beta_int` are estimated: sampling mode uses MCMC posterior means, while optimizing mode uses the MAP estimates. In both modes, `se_beta`, `se_beta_int`, and `se_total` are computed from the inverse negative Hessian.

```{r two-sample-optimizing, eval=FALSE}
result_2sample_opt <- int2MR(data_list_2sample = example_2sample_data,
                 model_type = "2sample",
                 estimation = "optimizing",
                 prior_inv_gamma_shape = 0.1,
                 prior_inv_gamma_scale = 0.1)

result_2sample_opt$result_2sample
```

## Example 2: Three-Sample Data

In this example, we perform an analysis on simulated three-sample data. The configuration is similar to the two-sample analysis, except the `data_list_3sample` argument is used instead of `data_list_2sample` and `model_type` is set to "3sample".

```{r three-sample-sampling, eval=FALSE}
result_3sample <- int2MR(data_list_3sample = example_3sample_data,
                 model_type = "3sample",
                 prior_inv_gamma_shape = 0.1,
                 prior_inv_gamma_scale = 0.1,
                 chains = 2, iter = 10000, warmup = 2500,
                 adapt_delta = 0.95)

# Display the results for the three-sample analysis
result_3sample$result_3sample
```

The three-sample model can also be run with `estimation = "optimizing"` to use the same optimization-based estimation path.
